### 2023-02-25 abstract.tex
Updates from Overleaf

\begin{abstract}
Type systems typically only define the conditions under which an expression is well-typed, leaving ill-typed expressions formally meaningless. This approach is insufficient for developing language servers for modern programming environments, which are expected to sensibly localize type errors and continue to provide a variety of semantic services in their presence. Lacking formal foundations, error localization has largely become a matter of folklore in the community. Recent work on the Hazelnut calculus takes a step toward assigning formal meaning to erroneous programs, but it stops short of a comprehensive account. In particular,  (1) Hazelnut is able to reason around marked type inconsistencies, but not several other kinds of static errors, and (2) Hazelnut provides an insufficient account of the marking procedure itself, relying on a limited structure editor, or on the user, to correctly insert marks. This paper addresses these deficiencies, contributing the first comprehensive formal account of type error localization, called the marked lambda calculus. This contribution is independently useful as a core calculus for language servers. In addition, we deploy its total marking procedure to fix several instances of undefined behavior in the Hazelnut structure editor calculus, namely by allowing edits that introduce non-local errors. We mechanize the calculus in Agda and implement an extension in Hazel, a full-scale live functional programming environment previously based on Hazelnut.

The marked lambda calculus supports local type inference, i.e. it is bidirectionally typed. Constraint-based type inference schemes (e.g. Hindley-Milner) are notorious for complicating error localization. We approach the problem of  constraint-based type inference in a novel manner: we deploy constraint solving as a layer atop a gradual bidirectionally typed core, where it infers type hole fillings. This system is able to elegantly handle situations where the system of constraints cannot be solved, because the unification algorithm we develop localizes errors arising from inconsistent constraints to individual type holes, rather than attempting to localize them to particular expressions, which is fundamentally \emph{ad hoc}. The user can resolve these errors by selecting from suggested partial solutions, or leave the errors unresolved, in which case they operate as the unknown type from gradual typing. We formally specify this approach and implement it in Hazel. 
\end{abstract}

>>>>

\begin{abstract}
Type systems typically only define the conditions under which an expression is well-typed, leaving ill-typed expressions formally meaningless. This approach is insufficient for developing language servers for modern programming environments, which are expected to sensibly localize type errors and continue to provide a variety of semantic services in their presence. Lacking formal foundations, error localization has largely become a matter of folklore in the community. Recent work on the Hazelnut calculus takes a step toward assigning formal meaning to erroneous programs, but it stops short of a comprehensive account. In particular,  (1) Hazelnut is able to reason around marked type inconsistencies, but not several other kinds of static errors, and (2) Hazelnut provides an insufficient account of the marking procedure itself, relying on a limited structure editor, or on the user, to correctly insert marks.\todo{decenter Hazelnut?} This paper addresses these deficiencies, contributing the first comprehensive formal account of type error localization, called the marked lambda calculus. This contribution is independently useful as a core calculus for language servers. In addition, we deploy its total marking procedure to fix several instances of undefined behavior in the Hazelnut structure editor calculus, namely by allowing edits that introduce non-local errors. We mechanize the calculus in Agda and implement an extension in Hazel, a full-scale live functional programming environment previously based on Hazelnut.

The marked lambda calculus supports local type inference, i.e. it is bidirectionally typed. Constraint-based type inference schemes (e.g. Hindley-Milner) are notorious for complicating error localization. We approach the problem of  constraint-based type inference in a novel manner: we deploy constraint solving as a layer atop a gradual bidirectionally typed core, where it infers type hole fillings. This system is able to elegantly handle situations where the system of constraints cannot be solved, because the unification algorithm we develop localizes errors arising from inconsistent constraints to individual type holes, rather than attempting to localize them to one particular expression, which is fundamentally \emph{ad hoc}. The user can resolve these errors by selecting from suggested partial solutions, or leave the errors unresolved, in which case they operate as the unknown type from gradual typing. We formally specify this approach and implement it in Hazel. 
\end{abstract}

### 2023-02-25 calculus.tex
Updates from Overleaf

\section{Motivating Examples}
In this section, we motivate a formal calculus for bidirectional type error localization through a
number of examples.

% TODO Talk about idea that bidirectional typing makes error localization easy?

\subsection{Unbound Variables}
% A simple example of an error that may occur in programs is the use of an unbound variable.

\subsection{Inconsistent Types}
Often, the type an expression \emph{should} be is known. For example, in an expression such as
$\EPlus{\EMV_1}{\EMV_2}$, both $\EMV_1$ and $\EMV_2$ should have the type $\TNum$. In a
bidirectional system, $\EMV_1$ and $\EMV_2$ would be analysed against $\TNum$.

\subsection{Inconsistent Branches}

\section{Marking Calculus}
\label{sec:calculus}

The previous section introduced different error localization cases by example. Now, we formally
introduce a \emph{marking calculus} for bidirectional type error localization on a lambda calculus
extended with numbers, conditionals, and let bindings, given by a two-layer system:
%
\begin{itemize}
  \item \textbf{Unmarked expressions} ($\EMName$), which are the surface-level expression syntax.
    Types classify unmarked expressions according to a standard bidirectional gradual typing
    semantics. Note that a non-gradual system may be recovered straightforwardly by removing the
    unknown type and replacing type consistency with equality (discussed in
    \cref{sec:calculus-non-gradual}).

  \item \textbf{Marked expressions} ($\ECMName$), which mirror the structure of unmarked expressions
    but are extended with \textbf{error markings}. Every marked expression corresponds to an
    unmarked expression by \emph{mark erasure}.

  \item \textbf{Mark insertion}, which \textbf{marks} any unmarked expression into a marked
    expression, inserting error markings where appropriate.
\end{itemize}

In the interest of space, this overview focuses on the ``interesting'' cases; the whole system,
alongside the metatheorems discussed have been mechanized in the Agda proof assistant (discussed in
\cref{sec:calculus-agda}).

\subsection{Unmarked Expressions}
\label{sec:calculus-uexp}

\Cref{fig:calculus-syntax-uexp} gives the syntax of types, $\TMV$, and unmarked expressions, $\EMV$,
which extend the simply typed lambda calculus with $\TNum$ and $\TBool$ base types of numbers and
booleans. The number literal corresponding to the mathematical number $n$ is given by $\ENumMV$, and
there is a single addition operation on numeric expressions. $\ETrue$ and $\EFalse$ correspond to
the boolean values $\textsf{true}$ and $\textsf{false}$, and the branching expression
$\EIf{\EMV_1}{\EMV_2}{\EMV_3}$ gives an if-else form.

In addition, types are extended with an \emph{unknown type}, $\TUnknown$, which

\begin{figure}[htbp]
  \[\begin{array}{rrcl}
    \TMName  & \TMV  & \Coloneqq & \TUnknown \mid \TNum \mid \TBool \mid \TArrow{\TMV}{\TMV} \\
    \EMName  & \EMV  & \Coloneqq & x \mid \ELam{x}{\TMV}{\EMV} \mid \EAp{\EMV}{\EMV} \mid \ELet{x}{\EMV}{\EMV} \\
             &       & \mid         & \ENumMV \mid \EPlus{\EMV}{\EMV}
                       \mid           \ETrue \mid \EFalse \mid \EIf{\EMV}{\EMV}{\EMV}
  \end{array}\]
  \caption{Syntax of types and unmarked expressions.}
  \label{fig:calculus-syntax-uexp}
\end{figure}

Unmarked expressions are governed by a standard bidirectional gradual type system, as given in
\Cref{fig:calculus-typing-uexp}.

% TODO Not sure how explicitly we want to spell out the type system, in lieu of background
% section on bidirectional typing?

% \input{figs/calculus-type-judgments.tex}
% \input{figs/calculus-typing-uexp.tex}

\subsection{Marked Expressions}
\label{sec:calculus-mexp}

Marked expressions, $\ECMV$, extend the syntax of unmarked expressions with explicit \emph{error
markings}. As given in the syntax definition of \cref{fig:calculus-syntax-mexp}, there are three
such forms, which mirror the examples above:
%
\begin{itemize}
  \item $\ECUnbound{x}$, denoting the usage of an unbound variable.

  \item $\ECInconType{\ECMV}$, which occurs when the synthesized type of $\ECMV$ is inconsistent
    with the required type.

  \item $\ECInconBr{\ECMV_1}{\ECMV_2}{\ECMV_3}$, which occurs when the types of the two branches
    $\ECMV_2$ and $\ECMV_3$ are inconsistent.
\end{itemize}
%
Additional forms may arise as required by the language.

\begin{figure}[htbp]
  \[\begin{array}{rrcl}
    \ECMName & \ECMV & \Coloneqq & x \mid \ECLam{x}{\TMV}{\ECMV} \mid \ECAp{\ECMV}{\ECMV} \mid \ECLet{x}{\ECMV}{\ECMV} \\
             &       & \mid         & \ECNumMV \mid \ECPlus{\ECMV}{\ECMV}
                       \mid           \ECTrue \mid \ECFalse \mid \ECIf{\ECMV}{\ECMV}{\ECMV} \\
             &       & \mid         & \ECUnbound{x} \mid \ECInconType{\ECMV} \mid \ECInconBr{\ECMV}{\ECMV}{\ECMV}
  \end{array}\]
  \caption{Syntax of marked expressions}
  \label{fig:calculus-syntax-mexp}
\end{figure}

Marked expressions are also typed by a bidirectional system, which, for most forms, closely mirrors
the rules for corresponding unmarked forms. \Cref{fig:calculus-typing-mexp} contains a selection of
the ``interesting'' cases, particularly those concerned with error markings.

\input{figs/calculus-typing-mexp.tex}

% TODO Not really sure how to explain the syntax and typing of marked expressions in an intuitive
% way in relation to unmarked expressions.

For unbound variables $\ECUnbound{x}$, nothing may be said about their types; hence, they synthesize
the unknown type. Inconsistent branch forms behave similarly, as inconsistent types have no join.

There are three rules concerned with the typing of lambda abstractions. MALam1, applies to
$\ECLam{x}{\TMV}{\ECMV}$ if the type $\TMV_3$ analysed against is a matched arrow type
$\TArrow{\TMV_1}{\TMV_2}$ and input type $\TMV_1$ is consistent with the binding ascription $\TMV$.
This corresponds to the lambda rule for unmarked expressions. However, we are additionally concerned
with the cases when $\TMV_3$ is not a matched arrow type or $\TMV_1$ is inconsistent with $\TMV$.
These are described by MALam2 and MALam3.

Alongside the standard subsumption rule, in which an expression that synthesizes a type may be
analysed against any consistent type, MAInconsistentTypes governs most occurrences of the
inconsistent types marking: $\ECInconType{\ECMV}$ analyzes against any type as long as $\ECMV$
synthesizes an inconsistent type. In both cases, we stipulate that subsumption may be applied to
$\ECMV$, written $\subsumable{\ECMV}$. In particular, lambda abstractions and if-else forms are
\emph{not} subsumable, formalizing these rules as ones of ``last resort'', preserving determinism of
typing.

Similarly to lambda abstractions, two rules govern application. In the unmarked syntax, the
application $\EAp{\EMV_1}{\EMV_2}$ was well-typed if the type $\TMV$ synthesized by $\EMV_1$ was a
matched arrow type. This corresponds to the MSAp1 rule for marked expressions. In the case that
$\TMV$ is not a matched arrow type, we require an inconsistent types marking around the

\subsection{Mark Insertion}

\subsection{Well-Formedness}
\label{sec:calculus-wellformedness}

\subsection{Alternative Formulations}

\subsection{Non-Gradual System}
\label{sec:calculus-non-gradual}

\subsection{Mechanization}
\label{sec:calculus-agda}

The metatheory of the calculus has been fully mechanized in the Agda proof assistant [citation]. The
mechanization's documentation contains more details regarding technical decisions made therein.

>>>>

\section{Marked Lambda Calculus}
\label{sec:calculus}
